---
title: "nmrdata: Example 1D Proton NMR Data"
author: "Torben Kimhofer"
date: "`r BiocStyle::doc_date()`"
package: nmrdata
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{nmrdata: Example 1D Proton NMR Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, message = FALSE, warning = FALSE)
```

# Overview

The **nmrdata** package provides example one-dimensional (^1H) NMR spectra of murine
urine samples collected before and after Roux-en-Y gastric bypass. A sham-surgery group
is included as a negative control [@li2011gut].

The package ships:

- a **processed dataset** (R list) for quick examples;
- a minimal set of **raw Bruker NMR experiment folders** for I/O demonstrations.

These data are intended for teaching and for demonstrating the companion package
`metabom8`.

```{r show-metadata}
meta_path <- system.file("extdata", "metadata.csv", package = "nmrdata")
if (file.exists(meta_path)) {
  meta <- utils::read.csv(meta_path)
  head(meta[c("Title","RDataPath")])
}
```

# Dataset summary

- Instrument: Bruker Avance III 600 MHz  
- Samples: 67 rat urine specimens  
- Processing: Baseline corrected, res. water signal excised, dilution normalised (probabilistic quotient normalisation)
- Provided as an R list with:
  - `X.pqn`: matrix of PQN-processed spectra
  - `ppm`: chemical shift axis (ppm)
  - `an`: sample annotation (e.g., class membership)
  - `meta`: acquisition/processing parameters (TopSpin; `a*`, `p*`)

# Installation

```{r, install-lib}
# Install from Bioconductor
# if (!require("BiocManager")) install.packages("BiocManager")
# BiocManager::install("nmrdata")
```


# Using the processed dataset (`.RData`)

> Note: During review the data are hosted on Zenodo; code below is **not** executed
> on Bioconductor’s build machines.

```{r load-processed, eval=FALSE}
library(nmrdata)

# load pre-processed data (returns a named list)
bariatric <- nmrdata:::loadBariatric(as = "list", quiet = TRUE)

str(bariatric, max.level = 1)

# visualise the first NMR spectrum
plot(bariatric$ppm, bariatric$X.pqn[1, ], type = "l",
     xlab = "Chemical shift (ppm)", ylab = "Intensity")
```

# Using the raw Bruker experiments (folder dataset)

```{r raw-dir, eval=FALSE}
# download once, unpack once; returns the directory path
exp_dir <- nmrdata:::getRawExpDir(quiet = TRUE)

# show top-level contents
list.files(exp_dir)
```

If you have `metabom8` installed, you can import/process the raw experiments:

```{r metabom8, eval=TRUE}
# library(metabom8)
# 
# # import Bruker 1D NMR spectra
# res <- read1d_proc(exp_dir, exp_type = list(pulprog = "noesypr1d"))
# 
# # plot first spectrum
# spec(res$X[1, ], res$ppm)
```

# Caching and reproducibility

Data are cached under a per-user directory so repeated calls don’t re-download:

```{r cache-info, eval=FALSE}
# where the archive and unpacked folder live
dir <- getRawExpDir(quiet = TRUE)
print(dir)
```

# Package version

```{r}
packageVersion("nmrdata")
```

# Session info

```{r}
sessionInfo()
```
