---
title: "Tutorial I"
#output: flexdashboard::flex_dashboard
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
  
#library(kableExtra)
library(DT)
library(htmlTable)
#library(formattable)
library(dplyr)
#library(flexdashboard)
```

# Data import, quality control and spectral pre-processing

&nbsp;

This tutorial runs through a typical data import and pre-processing workflow for Nuclear Magnetic Resonance (NMR) spectroscopy - based metabolic profilig with the *metabom8* R package. 

Let's get started by installing and loading the required R libraries:

```{r load-pack, fig.show='hold'}
# obtain NMR training data from GitHub
#install.packages('remotes')
#remotes::install_github('tkimhofer/nmrdata')

# install package for NMR data pre-processing and multivariate stats
#remotes::install_github('tkimhofer/metabom8')

# load packages
library(nmrdata)
library(metabom8)
```

&nbsp;


## Data

This tutorial uses ^1^H NMR data acquired from murine urine samples, available in the *nmrdata* package (www.github.com/tkimhofer). Urine samples were collected in a time-resolved fashion with a single collection before and multiple collections after bariatric surgery was performed. 1D ^1^H NMR spectra were acquired using standard 1D NMR experiments performed on a 600 MHz Bruker Avance III spectrometer, equipped with a 5 mm triple resonance (TXI) probe operating at 300 K. Further information on sample collection, processing and data acquisition can be found in the respective primary research article from Jia Li *et al.*^[Li, Jia V., *et al.* (2011) Metabolic surgery profoundly influences gut microbial-host metabolic cross-talk. *Gut* 60.9, 1214-1223.]

&nbsp;



## Import of NMR spectra and metadata
The **`read1d()`** function reads Bruker NMR spectra into the R workspace. The function requires at minimum two input arguments, that is: 1. *path*: the folder directory that encloses the NMR experiments and 2. *exp_type*: list of spectrometer parameter values to sub-select experiment types. In the present case, we will import all experiments that were performed with the pulse sequence program (PULPROG) 'noesypr1d':

```{r read-in, fig.show='hold'}
# define directory path of data encolsing folder
path=system.file("extdata/", package = "nmrdata")
# import 1D MRS data
read1d(path, exp_type=list(PULPROG='noesypr1d'))
```


Three spectral variables are routinely imported into the R workspace upon execution of the read1d() function:
```{r echo=FALSE, out.width='30%', include = T}
# create table
df=data.frame('Variable'=c('X', 'ppm', 'meta'), 
             'Description'=c('Matrix of NMR data matrix (rows = spectra, columns = spectral variables)', 
                             'Array of chemical shift information (ppm)', 
                             'Dataframe of spectrometer metadata'))

datatable(df, rownames = FALSE, options = list(dom = 't'), colnames = c('Variable Name' = 1))

# 
# £, align=c('c', 'l')) %>%
#   kable_styling(full_width = F) %>%
#   column_spec(1, bold = T, border_right = F) %>%
#   column_spec(2, width = "30em")

```



Rownames of *X* indicate the folder location of individual experiments, which can further be used to match sample annotation data (e.g., group-memberships). The dataframe *meta* is row-matched to the NMR data matrix (*X*) and contains metadata relating to acquisition and processing parameters, including the acquistion date. See `?read1d()` for more information. 


&nbsp;

## Visualisation of NMR spectra

###  Basic graphics
There are two low-level plotting functions for the visualisation of NMR spectra: **`spec()`** and **`matspec()`**. These functions can be used to plot a single spectrum or an overlay of multiple spectra, respectively. Both fuction can generate a static or an interactive plot, depending on the fucntion parameter *interactive*). The former is less compuatation expensive and exectutes faster. The latter takes longer to run but allows to to dynamically zoom-in to spectral areas. The data set in this tutorial is fairly small, so we will use the interactive version, which is the default plotting mode.

```{r spec overlay I, fig.show='hold', fig.width = 7.2, fig.height = 4}

# use 'spec' to plot a single pectrum, e.g., in row position 15:
#spec(ppm, X[15,], shift = range(ppm))

# use 'matspec' to overlay spectra, in ppm range 0 to 10
matspec(X, ppm, shift = range(ppm))
```

From this overview we can see that the spectral width ranges from -5 to approximately 15 ppm with the residual water signal resonating around 4.8 ppm. All ^1^H NMR signals are situated within the ppm range of 0 and 10, therefore both ends can be capped and this will be illustrated further below.

### ggplot2 

There are different higher-level plotting functions available that allow a more comprehensive spectral visualisation of individual peaks or spectral ranges, which are mainly used to create publication-quality images. One of these functions is **`specOverlay()`**, which is comparable to **`matspec()`** shown above, but enables to create different subplots (aka facetting) and to include colour scales and linetypes in a straightforward way. 


### Plotting function parameter *an*
Point/line colour, labels and sub-division (facetting) of plots create insight. In ggplot-based plotting functions of *metabom8*, these formatting options are provided as a list with the function argument *an* . The order in which the plotting parameters are provided matters and is always the same: `an=list(facet, colour, lineype)`, whereas each list element is a vector with as many elements as there are spectra (=rows) in *X*.

For illustration purposes, let's plot the TSP singlet located in the ppm range of -0.05 to 0.05 and add information on acquisition paramters that can be found in the *meta* dataframe: the different panels (facetting) show different experiment types, the colour maps to the date stamp of the acquisition (*a_Date*) and the linetype indicates the NMR pulse program (*a_PULPROG*):

```{r spec overlay II, fig.show='hold', fig.width = 7, fig.height = 5}
# plot TSP signal
specOverlay(X, ppm, shift=c(-0.05,0.05), 
    an=list( 'Facet'=meta$a_EXP, # facet
             'Date'=meta$a_DATE, # colour
             'Pulse Program'=meta$a_PULPROG) # linetype
    ) # linetype

```

The plot above shows that two different experiment types were performed (i.e., two different panels), all based on the pulse program *\<noesypr1d\>*, which a Bruker term for a standard 90˚ pulse experiment with water pre-saturation. Experiment types labelled *\<\>* are calibration experiements and were performed before type *\<JL-noe1D\>* (see Date information, i.e., run order).


## Spectral referencing

Spectral referencing refers to horizontally shifting an entire spectrum unitl a signal of a standard compound reaches a defined chemical shift position. In urine metabolic profiling, the standard compound is TSP^[3-(trimethylsilyl)-2,2′,3,3′-tetradeuteropropionic acid] (usually added to the sample buffer), which gives rise to a singlet that is defined to resonate at 0 ppm.^[Dona, A.C., *et al.* (2014) Precision high-throughput proton NMR spectroscopy of human urine, serum, and plasma for large-scale metabolic phenotyping. *Analytical Chemistry*. 86.19. 9887-94.]

We can reference the urine spectra using the **`calibrate()`** function, as shown with the following code:
```{r calibration, fig.show='hold', fig.width = 7.2, fig.height = 6}
# perform TSP calibration
X.cal=calibrate(X, ppm, type='tsp')

# plot TSP after calibration
matspec(X.cal, ppm, shift=c(-0.1,0.1))
```

Now you can see that for all spectra, the TSP peak apex is centered at a chemical shift of zero ppm (that was not the case before). 



&nbsp;

## Filtering based on spectrometer parameters
For statistical analysis we are only interested in *\<JL-noe1d\>* experiment type, as the other calibration experiments are not suitable for quantitative analysis. To exclude the latter, execute the following R code which also re-runs the previous plotting command:

```{r filter, fig.show='hold', fig.width = 7.2, fig.height = 4}
# Exclude calibration experiments
idx=grep('noe1d', meta$a_EXP) 
X.cal=X.cal[idx,]
meta=meta[idx,]

# plot TSP signal
specOverlay(X.cal, ppm, shift=c(-0.05,0.05), 
    an=list( 'Facet'=meta$a_EXP, # facet
             'Receiver Gain'=meta$a_RG, # colour
             'Pulse Program'=meta$a_PULPROG) # linetype
    ) # linetype
```


&nbsp;


## Assessment of spectral quality

In metabolic phenotyping and in any other field where multiple NMR spectra are compared quantitatively, the quality of spectra is of particular importance. Basic spectral quality assessement includes visual inspection of the water suppresion quality, peak symmetry, spectral line widths, baseline level and stability as well as the average signal to noise (SN) ratio.

Visual assessment of spectral quality:

```{r spec qc I, fig.show='hold', fig.width = 7.2, fig.height = 4}
# calculate quality control measures
matspec(X.cal, ppm)
  
```

&nbsp;


## Excision of signals

Further downstream analysis requires the excision of chemical shift regions bearing no biological or non-quantitative information. In urinary NMR analyses this includes the TSP signal, the residual water and urea signal as well as ppm regions where there is no signal but only noise.

The function **`get.idx()`** can be used to obtain indices of the desired shift range from the ppm vector. These indices can then be further used to exclude the relevant columns in the NMR matrix and ppm vector. This is illustrated in the following code snippet.

```{r excision, fig.show='hold', fig.width = 7.2, fig.height = 4}
# Indexing TSP region and upfield noise...
idx.TSP=get.idx(range=c(min(ppm), 0.5), ppm)
# ... water region...
idx.water=get.idx(range=c(4.6, 5), ppm)
# ... as well as downfield noise regions
idx.noise=get.idx(range=c(9.5, max(ppm)), ppm)

idx.rm=c(idx.TSP, idx.water, idx.noise)

# Exision of TSP, res. water and noise regions
X.cut=X.cal[,-idx.rm]
ppm=ppm[-idx.rm]
```

&nbsp;


## Baseline correction

In 1D 1H NMR spectroscopy using simple pulse sequences, broad resonanes originating from macromolecules lead to increased spectral baselines. Baseline differences across spectra complicates the analysis and the the removal of often leads to more accurate results. In MetaboMate, a non-linear baseline correction can be performed with the **`bline()`** function. See the exmpale below for its input arguments and **`?bline`** for methodological information.

```{r baseline, fig.show='hold', fig.width = 7.2, fig.height = 4}
# Baseline correction
X.bl=bline(X.cut)
# visual assessment
spec(X.cut[1,], ppm, shift = range(ppm))
spec(X.bl[1,], ppm, shift = range(ppm), add=T)

```

&nbsp;


## Spectral normalisation

Depending on the sample type, spectra may require normalisation prior to statistical analysis. For example, urine dilutions vary across samples, perhaps due to the uptake of different amounts of water. Normalisation methods can account for these kind of systematic differences in spectral intensities.

There are several normalisation methods avaiable. Among the most commonly applied ones are Total Area (TA) normalisation and Probablistic Quotient Normalisation (PQN).^[Dieterly, F., \emph{et al.} (2006), Probabilistic Quotient Normalization as Robust Method to Account for Dilution of Complex Biological Mixtures. Application in 1H NMR Metabonomics, \emph{Analytical Chemistry}, 78.3, 4281-90] In this tutorial, we normalise the spectra using both methods and compare the results. Therefore, the functions **`pqn()`** and **`totaArea()`** are called, both returning the normalised NMR matrix. The additional input argument *add.DilF* indicates if the calculated dilution factors should be exported to the R workspace. Having a look at the calculated dilution factors can be very informative. For example, if pooled quality control samples were periodically included in the study run, then these should have very comparable dilution factors. The **add.DilF** argument specifies a user-given variable name and a vector variable named like this (containing the dilution factors) is automatically added to the R workspace after calling the function. Here's an example: 

```{r normalisation, fig.show='hold', fig.width = 7.2, fig.height = 4}
# PQN normalisation
X.pqn=pqn(X.bl, bin=NULL, add_DilF = 'dilF.pqn')

# Visualising the pqn dilution coefficient / scaling factor for each spectrum
hist(dilF.pqn, xlab='PQN calculated dilution factor')
abline(a = c(0,1), col='green')
```


The final step in this preprocessing tutorial is the visual inspection of the pre-processed NMR spectra: 

```{r visal check, fig.show='hold', fig.width = 7.2, fig.height = 4}
matspec(X.pqn, ppm, shift = range(ppm), interactive=T)
```

&nbsp;


# Summary and further steps

1D NMR spectra were imported, quality checked and pre-processed which included referencing to TSP, baseline correction, excision of spectral areas bearing no quantitative biological information. Urinary spectra were normalised with PQN to account for different sample dilutions. 

The pre-processed spectra can now be statistical interrogated, e.g., using multivariate methods Prinicpal Component Analysis (PCA) and Orthogoanl-Partial-Least Squares (O-PLS). You can find more information on this in the vignette **Multivariate Statistical Analysis** of the *nmrdata* package.
